A Machine Profile Summary

### 2022-06-13 MachineSummary/MachineSummary.Rmd
JN first major edit of MachineSummary

# Providing a characterization of a particular computing environment for R calculations

If we want to compare results computed on different machines, we need a way to provide 
measures and identifiers of the particular computing environment at hand. This write-up
attempts to summarize some possibilities available in the R statistical software and language.
Note that one of our interests is to have tools that work in R **across different platforms**.
That is, we want to be able to call an R function to get machine and computing environment
characterizations regardless of the particular platform which we are currently using.
Moreover, we want to have a succinct output of such information.

# Some possible information desired

There are multiple ways to get information of the type useful for characterizing the
computing infrastructure used for a particular test or timing. However, we need to
ensure that identifiers of particular pieces of information are unique, and that 
common identifiers really do specify identical, or at least equivalent, information. 
Otherwise, new names for some pieces of our set of capability information are needed.

@@

When we wish to analyze performance results, it is extremely helpful to have 
information on when the calculations were performed. 
Most computations should have a time/date stamp that is unlikely to be confused with any
other, at least in combination with a machine name and/or other tags. Here is one
suggestion based on the R `Sys.time()` function.

```{r tdstamp}
#get a timestamp
tsstr <- format(Sys.time(), "%Y/%m/%d|%H:%M") # tsstr == time stamp string in form YYYYmmddHHMM
cat("Date and Time stamp:", tsstr, "\n")
```

## Identifier for the computing software environment

@@

- a machine name. While most systems (and Linux in particular) offer to let the user
  provide a machine name, there are generally defaults that many people accept, and
  these are often uninformative and may be non-unique. We note that VirtualBox with
  a Windows 10 guest machine gave the name DESKTOP-HF4CKVA. Where this name was 
  generated we are not sure. Settings / System allows "rename this PC".

@@

- compiler or interpreter version. For our needs, it is obvious that the R version
  will be important. However, if any code is compiled or linked to libraries, we
  would like to know that. In particular, versions of compilers (e.g., gfortran, 
  gcc) or BLAS or LAPACK libraries used will affect performance. Linux generally
  displays the BLAS and LAPACK **filenames** in `sessionInfo()`, but to get the
  version information, one needs to dig deeper, for example, using operating
  system commands. For the present, we will omit such extra information unless
  we can extract it easily within R.

@@

Clearly it is important to identify which software packages are active, and the
R `sessionInfo()` command is useful for this. However, there may be additional 
features that should be mentioned if the user has modified the software in any
way.

@@

@@

An important tool for acquiring machine information is the CRAN package
`benchmarkme` (@benchmarkme). We found a minor bug in the `get_ram()`
function of this package for some Windows 10 versions. It has been 
corrected in the latest version 1.0.8, with contributions from one of the authors (AB).

<!-- on https://github.com/csgillespie/benchmarkme, but at the -->
<!-- time of writing is not yet on the CRAN repository. ??Do we wish to comment on the -->
<!-- "GB" issue? -->

@@

- uses `proc.time()` for timings
- has single core and parallel timings
- ?? does it do GPUs
- provides far too much for most needs i.e., a succinct summary of
  how capable a computing system should be

```{r benchmarkmex}
library(benchmarkmeData)
library(benchmarkme)
ls(package:benchmarkmeData)
ls(package:benchmarkme)
lsf.str("package:benchmarkme")
lsf.str("package:benchmarkmeData")
## This next line takes a lot of time to run, so is commented out here
# benchmark_std()
get_byte_compiler()
gla <- get_linear_algebra()
gsd <- get_sys_details()
```

## Calculating the machine precision

Until the IEEE 754 standard (?? @institute1985ieee) became available in the 1980s, 
computers used multiple forms of floating point arithmetic. With
some ingenuity, it is possible to detect many aspects of the 
floating point properties.
This is a port(??part) of Mike Malcolm's ENVRON to R. ??ref. It computes the 
machine precision, radix and number of radix digits.

?? mention Kahan's Paranoia

## Simple performance examples

There are many performance examples used for testing the speed of computers. A 
well-known example is the @linbench03 LINPACK test used for ranking supercomputers.
These generally employ linear algebra calculations. The example below, by
contrast, focuses on the special function computations. Variations on this
simple performance test have been used by one of the authors for nearly half
a century. Note the use of the `microbenchmark` timing package (@microbench13).

```{r loopesc}
loopesc <- function(nn){
  ss <- 0
  for (i in 1:nn) {
    xx <- exp(sin(cos(1.0*i)))
    ss <- ss + xx
  }
  xx
}
require("microbenchmark")
nvals<-c(1000, 10000, 100000, 1000000)
tvals<-rep(NA, length(nvals))
i<-0
for (nn in nvals) {
  i <- i + 1
  cat(nn, "\n")
  tt <- microbenchmark(loopesc(nn), unit = 'us')
  print(tt)
  tvals[i] <- mean(tt$time)
}
plot(log10(nvals), log(tvals))

```

## Issues relating to random number generation

## Issues relating to compilation of R

- LAPACK
- BLAS
- others?

# Tools for accessing and clearing environments and dataframes

?? should this be here -- probably to list things in the workspace

```
sys.frame()
sys.frames()
sys.status()
sys.on.exit()
sys.parents()
sys.calls()
```
# Choices

For use in recording tests of R functions and packages for optimization and nonlinear 
least squares, it seems that the `benchmarkme` function `get_sys_details()`
provides more than sufficient information for our needs.

From the above discussion, the following offers a possible compact solution. Users may
wish to modify this to their own particular needs.

@@

Note that we are reading from the file `MachID.R` so that this file is usable elsewhere
and the current document does not become out of sync with the working MachID.

>>>>

# Abstract

Many statistical calculations draw on system resources of memory, storage, arithmetic,
instructional processing and communications. A given computational environment will
have a particular collection of such resources. With workers now commonly working 
separately and sharing their work virtually, we should consider how to characterize
the systems under which calculations are carried out in order that we can diagnose
differences in results or performance of the computing methods. Our objective here
is to consider how a succinct labeling may provide a **machine profile summary**
to assist such analysis and comparisons. To focus our work, we limit our attention
to R as the software environment, but the ideas are more generally applicable.

# Characterizing a particular computing environment for R

In 2021, we undertook an investigation of tools in R for nonlinear least squares
calculations under the Google Summer of Code initiative. In the process, we recognized 
that we would be comparing calculations
performed on **different computing environments** at our disposal. Some of these were
based on different hardware, while others were different combinations of given hardware
with alternative operating software. In one case a machine failed and had to be replaced,
underlining the need to be able to compare results from the defunct system with its
replacements.

A useful context for our ideas is provided by the Wikipedia article on Computer Performance
(https://en.wikipedia.org/wiki/Computer_performance). We seek a way to provide succinct
measures and identifiers of the particular computing environment at hand, leading to a
**profile** for a given computing environment for which we will provide a 
short but meaningful label unique within our project context. This article
considers some possibilities available in the R statistical software and language
that assist this effort, but the ideas are extensible to other programming systems.

# Desired information

There may be multiple terms used to describe some of the information we want, as
well as different interpretations of some of the units of measure. For example
"Gigabytes", or GB, may be $1000^3$ or $1024^3$ bytes. ?? AB? Is that correct?

@@

For tracking work, it is important to have 
information on when the calculations were performed. 
Most computations should have a time/date stamp that is unlikely to be confused with any
other, at least in combination with a machine name and/or other tags. Here is one
suggestion based on the R `Sys.time()` function. Moreover, as we were working on opposite
sides of the globe, it could be important to either specify time zone or to work with a
single "clock" such as UTC (ref?? or expansion?)

```{r tdstamp}
#get a timestamp
tsstr <- format(Sys.time(), "%Y/%m/%d|%H:%M") # tsstr == time stamp string in form YYYYmmddHHMM
cat("Date and Time stamp:", tsstr, "\n")
# and here is a way to get the UTC time
tutc<-as.POSIXlt(Sys.time(), tz = "UTC")
print(tutc)
utctstmp<-format(tutc, "%Y/%m/%d|%H:%M")
print(utctstmp)
```

## Computing software environment

@@

- a machine name. While most systems (and Linux in particular) offer to let the user
  provide a machine name, there are generally defaults that many people accept, and
  these are often uninformative and may be non-unique. We note that VirtualBox with
  a Windows 10 guest machine gave the name DESKTOP-HF4CKVA. Where this name was 
  generated we are not sure. Settings / System allows "rename this PC". Clearly 
  we may want to extend the name if particular software is involved, as follows.

@@

- compiler or interpreter version. For our needs, it is obvious that the R version
  will be important. However, if any code is compiled or linked to libraries, we
  would like to know that. In particular, versions of compilers (e.g., gfortran, 
  gcc) or BLAS or LAPACK libraries used will affect performance. Linux generally
  displays the BLAS and LAPACK **filenames** in `sessionInfo()`, but to get the
  version information, one needs to dig deeper, for example, using operating
  system commands. 
  <!-- ??For the present, we will omit such extra information unless -->
  <!-- we can extract it easily within R. -->

@@

Clearly it is important to identify which software packages are active, and the
R `sessionInfo()` command is useful for this. However, there may be additional 
features that should be mentioned if the user has modified the software in any
way, or is using customized packages.

@@

## Machine precision and .Machine

Until the IEEE 754 standard (@institute1985ieee) became available in the 1980s, 
computers used multiple forms of floating point arithmetic. With
some ingenuity, it is possible to detect many aspects of the 
floating point properties. In particular, we may be interested in the
**machine precision**, that is, the smallest number `eps` such that 
`1 + eps` is greater than `1`. R performs computations in IEEE "double"
precision, and provides the R variable `.Machine` that has " information on 
the numerical characteristics of the machine R is running on", including
the machine precision as `.Machine$double.eps`.

The information in `.Machine` is documented as based on @Cody1988Alg665, which is 
an extension of Mike Malcolm's ENVRON (@Malcolm72). A port of this to R is
presented below. It computes the 
machine precision, radix and number of radix digits. A more detailed 
tool was W. Kahan's PARANOIA (@Karpinski85). The rise of IEEE arithmetic has
set aside the routine use of such tools, but recent innovations in so-called
"half-precision", of which there is more than one variety, may result in 
renewed interest. See https://en.wikipedia.org/wiki/Half-precision_floating-point_format.

This R function returns information on the operating system and the generic type of the 
processor.

@@

The CRAN package `benchmarkme` (@benchmarkme) provides a wide range of
information on system in which it is running, indeed a superset of 
previous tools mentioned. It is primarily designed to 
benchmark the capabilities of a particular computing system, so highly appropriate
to our current effort, though its output is not succinct. Note that we found a 
minor bug in the `get_ram()`
function of this package for some Windows 10 versions. Our thanks for 
Colin Gillespie, the author, for making  
corrections in version 1.0.8, based on communications with one of us (AB).

@@

?? AB: I found the plots of the benchmarks difficult to read. Which end of the
plots are "better"?

<!-- - uses `proc.time()` for timings -->
<!-- - has single core and parallel timings -->
<!-- - ?? does it do GPUs -->
<!-- - provides far too much for most needs i.e., a succinct summary of -->
<!--   how capable a computing system should be -->

<!--  ```{r benchmarkmex}  -->
<!-- # library(benchmarkmeData) -->
<!-- # library(benchmarkme) -->
<!-- # ls(package:benchmarkmeData) -->
<!-- # ls(package:benchmarkme) -->
<!-- # lsf.str("package:benchmarkme") # list functions in -->
<!-- # lsf.str("package:benchmarkmeData") -->
<!-- ## This next line takes a lot of time to run, so is commented out here -->
<!-- # benchmark_std() -->
<!-- # get_byte_compiler() -->
<!-- # gla <- get_linear_algebra() -->
<!-- # gsd <- get_sys_details() -->
<!--  ``` -->

## Some other performance examples

There are many performance examples used for testing the speed of computers. A 
well-known example is the @linbench03 LINPACK test used for ranking supercomputers.
These generally employ linear algebra calculations. The example below, by
contrast, focuses on the special function computations. Variations on this
simple performance test have been used by one of the authors for nearly half
a century. We used the `microbenchmark` timing package (@microbench13),
which allows us to see the variation in timings, which can be the result of
other activities and processes in our computing systems.

Note that it appears that the first cycle of the loop is very slow. ?? is this
byte-compiling?

```{r loopesc}
loopesc <- function(nn){
  ss <- 0
  for (i in 1:nn) {
    xx <- exp(sin(cos(1.0*i)))
    ss <- ss + xx
  }
  xx
}

require("microbenchmark")
# nvals<-c(1000, 10000, 100000, 1000000)
# tvals<-rep(NA, length(nvals))
# i<-0
# for (nn in nvals) {
#   i <- i + 1
#   cat(nn, "\n")
  nn <- 10000
  library(compiler)
  le <- cmpfun(loopesc)
  te <- microbenchmark(le(nn), unit='us')
  print(te)
  boxplot(te)
  tt <- microbenchmark(loopesc(nn), unit = 'us')
  print(tt)
  boxplot(tt)
#  ht<-hist(tt$time)
#  plot(ht, main="Histogram of loopesc times")
  tt2 <- microbenchmark(loopesc(nn), unit = 'us', control=list(warmup=2))
  print(tt2)
  boxplot(tt2)

#  ht2<-hist(tt2$time)
#  plot(ht2, main="Histogram of loopesc times after 2 warmup cycles")
# }
# plot(log10(nvals), log(tvals))
```

## Missing tools

While we cannot be certain that they do not exist, there do not appear to be tools
readily available to allow us to characterize the following features of R computations
that might be relevant to understanding machine performance.

- Some users will want to know key information about (pseudo-)random number generation
  used in their calculations.
  We suspect that users will need to custom-code ways to get such information
  about their computations. For example, it is useful to have the specification
  of the generator and the seed(s) used. For parallel computations, it is important
  in some applications to use the same set of numbers for each thread, while in
  others we want to use separate segments of the stream of pseudo-random numbers. 
  From the specification of the generator, it should be possible to know the 
  period (how many numbers are generated before the sequence repeats) and 
  properties concerning correlations between generated numbers. However, in our
  experience, such information requires a lot of research to uncover.

- R allows for byte compilation of code, and there are several ways in which this
  can be done. `benchmarkme` returns some information about the general setting used,
  but details would need user intervention if locally written code is used.

- Compilation of code written in other programming languages and how the resultant
  routines are called will similarly require user attention. Fortunately, some 
  information is available for heavily used software such as the BLAS and LAPACK,
  for which `sessionInfo()` reports the libraries used. There are, however, several
  alternatives, some of which claim high-performance.

<!-- # Tools for accessing and clearing environments and dataframes -->

<!-- ?? should this be here -- probably to list things in the workspace? -->

<!-- ``` -->
<!-- sys.frame() -->
<!-- sys.frames() -->
<!-- sys.status() -->
<!-- sys.on.exit() -->
<!-- sys.parents() -->
<!-- sys.calls() -->
<!-- ``` -->

# A modest profile

For most uses in recording tests of R functions and packages for optimization and nonlinear 
least squares, which has been our focus, it seems that the `benchmarkme` function 
`get_sys_details()` provides more than sufficient information for our needs. 
The following script offers a possible compact solution. Users may
wish to modify this to their own particular needs.

@@

<!-- Note that we are reading from the file `MachID.R` so that this file is usable elsewhere -->
<!-- and the current document does not become out of sync with the working MachID. -->
